---
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

### Assigning values to objects
`<-` is the standard assignment operator. You will assign values to objects (e.g. `x`, `y`, and `j`) using this symbol. For example `x <- 3`  is an object with one element, `x` is assigned a value of 3.

`c` is a common function used for concatenation (think combine): 
```{r, eval = FALSE}
x <- c(3,7,11,19)
y <- c(1,1,1,1)
c(x,y)
y <- c(x,y,5)
```

There are a suite of functions to enter data in the console. The sequence function `seq`:
```{r, eval=FALSE}
y <- seq(from = 1, to = 10)
y <- seq(from = 1, to = 10, by = 2.5)
y <- seq(from = 1, to = 10, length.out = 22)
```

Use the function `rep` (repeat):
```{r, eval=FALSE}
y <- rep(x = 5, times = 4)
x.value <- c(2,3)
rep(x.value, times = 3)
```

### Indexing
Indexing is used for "interogation" and "assignment". Interogation means to examine the value of an object. Assigment means to assign the value of an object.
```{r, eval=FALSE}
x <- c(3,7,11,19)
# Assignment
x[c(2,3)] <- c(120,140)
# Interogation
x <- c(3,7,11,19)
x[3]
x[4]
```

Extend the utility of indexing: `[]` by combining with `c`. For example to get the second and third element in the variable "x"
```{r, eval=FALSE}
x <- c(3,7,11,19)
x[1]
x[2]
x[c(1,2)]
x[c(2,1)]
y <- c(3,2,1)
x[y]
```

```{r, eval=FALSE}
# Extend the index using a negative sign
y.vect <- c(1,2,3,4,5,6)
y.vect[-3]
```

Indexing practice

Number| Excercise
------| -------------------------------------------------------------------
1.    |Can you get a + to occur on your console?
2.    |How many different ways can you create a vector labled `q` containing two 3's and four 5's? Try some!
3.    |Assign a vector of four elements: 3,7,9 and 2 to `w`.
4.    |Assign the third element of `w` to `s`, where `s` is equal to 6. 
5.    |What is the length of a sequence that starts at 1.1, ends at 9.2, and has increments of 0.894?
6.    |What is the 3rd value of the sequence you created?
7.    |Use the summary function on the sequence object you created, what is returned?

### Arithmetic operators
Symbol| Operation
------| ---------
+     | addtion
-     | subtraction
*     | multiplication
/     | division
^     | exponentiation

PEMDAS applies when writing R-code
```{r, eval=FALSE}
# Arithmetic operators in action:
5 + 5 / 2
3 * 2^2
(3*2)^2 
# Arithmetic operators using objects:
z <- 5
w <- c(3,7,9,2)
s <- w[3]
z + s
```

Arithmetic operators practice
Using what you know about parenthesis and PEMDAS, in one line of code do each of the following:  

Number | Excercise
------ | -------------------------------------------------------------------
1.     |Assign the variable `x` to be a vector containing the values 5,5,6,2  
2.     |Assign the variable `y` to be a vector containing the values 3,3,1,7
3.     |Add `x` and `y`
4.     |Substract `y` from `x` 
5.     |Assign `d` as `y` divided by `x`
3.     |Multiply `z` by `s` then add five (`z` and `s` from last practice)
4.     |Add 5 to `z` then multilply by `s`
5.     |Take `z` to the fifth power and then add 2
6.     |Divide `s` by three, then add 33, then take that sum to the 0.5 power


### Logical operators
Symbol | Operation
------ | ---------
<      | less than
>      | greater than
<=     | less than or equal to
>=     | greater than or equal to
==     | exactly equal to

Logical operators are useful in indexing and subsetting data classes. They return `TRUE` and `FALSE` values.
```{r, eval=FALSE}
# logical operators examples
y <- c(2,3,4)
w <- c(3,7,9,2)
y > 1
w == 2
w < 4 | w >= 8
w > 4 & w <= 8
``` 

Combine logical operators with the function `which`
```{r, eval=FALSE}
# logical operators examples
y <- c(2,3,4)
y > 1
which(y > 1)
which(y < 3)
which(y <= 3)
y[which(y <= 3)] <- 0
``` 

Logical operators practice

Number | Excercise
------ | -------------------------------------------------------------------
1.     |Assign a vector `a` to have six elements: 1,0,8,11,12,20
2.     |Which elements in `a` are less than five
3.     |Which elements in `a` are greater than 13 or less than 10

### Dealing with missing values

+ In R, missing values are represented by "NA". 

+ Undefined values (like dividing by zero) are represented by "NaN", not a number. Often missing values are represented with numbers: -1, 99, -9999, etc. This is obviously a problem and should be avoided. You will likely need to use indexing prior using arithmetic operations to replace these values.

The "is.na" function and the "na.rm"" argument:

+ Sometimes we do not know whether there are missing values in our data. 

+ We can use the `is.na` function to test for missing values:

```{r, eval=FALSE}
# is.na tests for missing values
dat.1 <- c(-1,NA,1,1,-1)
is.na(dat.1)
which(is.na(dat.1))
```

We can use the logical `na.rm` argument to remove missing values from our data prior to executing the funciton:
```{r, eval=FALSE}
dat.1 <- c(-1,NA,1,1,-1)
mean(dat.1)
mean(dat.1, na.rm = T)
```

Sometimes we code no data as -1 and that can really screw things up. R does not know that -1 means "no data". However we can replace the -1 with NA. There are many ways to do this, but here is a simple way. 
```{r, eval=FALSE}
# In this example, -1 is coded as a missing data field. Think back to our logical arguments and subsetting exercises.
dat.2 <- c(2,-1,3,4,5)
dat.2 == -1
dat.2[dat.2 == -1]
dat.2[dat.2 == -1] <- NA 
```

```{r, eval=FALSE}
dat.1 <- c(2,NA,3,4,5)
# is.na tests for missing values
is.na(dat.1) 
# returns the element(s) number in the vector that is NA
which(is.na(dat.1))
```

### Data classes
```{r ,eval=FALSE}
# Create a numeric vector using the concatenate function
y.vect <- c(2,3,4)
y.vect
class(y.vect)
length(y.vect)
```

```{r ,eval=FALSE}
# Create a numeric vector using the replicate function
y.vect <- rep(x = 2, times = 4)
y.vect
class(y.vect)
```

```{r ,eval=FALSE}
# Create a numeric vector the sequence function
y.vect <- seq(from = 1, to = 12)
y.vect
class(y.vect)
```

```{r ,eval=FALSE}
# Create a numeric vector the sequence function
y.vect <- seq(from = 1, to = 12, by = 3.4)
y.vect
class(y.vect)
```

Class Data Frames
```{r ,eval=FALSE}
# Create a data frame using the "data.frame" function
site.name <- c(rep("Site.01",3),rep("Site.02",3))
density <- rep(x = 2.3, times = length(site.name))
abundance <- seq(from = 14.5, to = 19.8, length.out = length(site.name))
sampled. <- c(F, T, F, F, T, F)
y.data.frame <- data.frame(site.name, density, abundance, sampled.)
y.data.frame
class(y.data.frame)
```

Class Matrices
```{r, eval=FALSE}
# Use the matrix function
y.mat <- matrix(data = c(2,3,4,5,6,7), nrow = 2, ncol = 3, byrow = FALSE)
y.mat
class(y.mat)
length(y.mat)
dim(y.mat)
```

```{r ,eval=FALSE}
# Use the matrix function
y.mat <- matrix(data = c(2,3,4,5,6,7), nrow = 2, ncol = 3, byrow = TRUE)
y.mat
class(y.mat)
```

Class Lists
```{r, eval=FALSE}
# Create the list
y.list <- list(site.name, density, abundance)
y.list
class(y.list)
```

```{r, eval=FALSE}
# Create list with variable names
y.list <- list(site.name = site.name, density = density, abundance = abundance)
y.list
class(y.list)
length(y.list)
```

Class Logical
```{r ,eval=FALSE}
# Create a vector of logical elements
y.vect
y.vect == 3.2
y.log <- c(y.vect == 3.2)
class(y.log)
```
