---
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

Survey of some of the capabilities of base R statistics functionality

```{r, eval = FALSE}
require(stats)
require(datasets)
require(utils)
```

### Z-scores
```{r, eval = FALSE}
scores <- c(76,80,83,97,100)
?scale
scaled.scores <- scale(scores)
hist(scores)
scores.data.frame <- data.frame(scores,scaled.scores)
scores.data.frame
plot(x = scores.data.frame$scores, 
     y = scores.data.frame$scaled.scores, 
     type = "b",
     xlab = "Raw Test Scores",
     ylab = "Scaled Test Scores")
abline(h = 0)
mean(scores)
abline(v = mean(scores))
```

```{r, eval = FALSE}
z.scores <- c(scores - mean(scores))/sd(scores)
scores.data.frame <- data.frame(scores,scaled.scores,z.scores)
scores.data.frame
```

### Quantiles
Here is an example using the quantile() function:
```{r, eval = FALSE}
?quantile
nile.vect <- Nile
hist(nile.vect, main = "", xlab = "Nile river discharge volume")
quantile(x = nile.vect)
abline(v = quantile(x = nile.vect), col = "red", lwd = 2)
```

### Normality 

Shaprio-Wilk test of normality

Null hypothesis - data are normally distributed

```{r, eval = FALSE}
?shapiro.test
?rbeta
test.dat.1 <- rbeta(n = 500, shape1 = 0.5, shape2 = 5)
hist(x = test.dat.1)
# Test if these data are normally distributed
shapiro.test(x = test.dat.1)
```

```{r, eval = FALSE}
test.dat.2 <- rnorm(n = 500, mean = 0, sd = 1)
hist(x = test.dat.2)
# Test if these data are normally distributed
shapiro.test(x = test.dat.2)
```

### t-test
```{r, eval=FALSE}
head(ToothGrowth,15)
un.dose <- unique(ToothGrowth$dose)
ind.1 <- which(ToothGrowth$dose == un.dose[1])
ind.2 <- which(ToothGrowth$dose == un.dose[2])

t.test(ToothGrowth$len[ind.1], ToothGrowth$len[ind.2])
```

### Linear Models
lm() function can be used to create a simple regression model. 
The lm() function accepts a number of arguments: 

  + formula: describes the model
    + This is "YVAR ~ XVAR" 
    + YVAR is the dependent (predicted)
    + XVAR is the independent (predictor)

```{r, eval=FALSE}
# Load the UNM_Enroll.csv data file
plot(x = mtcars$wt, 
     y = mtcars$mpg,
     xlab = "Weight", 
     ylab = "mpg", 
     type = "p", 
     las = F)

model.object <- mtcars$mpg ~ mtcars$wt
model.object
lm(model.object)
```

+ The intercept is 37.285 and the coefficient for the unemployment rate is  -5.344  
+ Therefore, the complete regression equation is mpg = 37.285 +  -5.344  * Weight. 

```{r, eval=FALSE}
# NHST
anova(lm(model.object))
summary(lm(model.object))
pred.y.vals <- predict(lm(model.object))
plot(x = mtcars$wt, 
     y = mtcars$mpg,
     xlab = "Weight", 
     ylab = "mpg", 
     type = "p", 
     las = F)
lines(x = mtcars$wt, pred.y.vals)
```

### Analysis of Variance
```{r, eval=FALSE}
# ANOVAs are linear models with categorical predictors
head(ToothGrowth)
anova.model.object <- ToothGrowth$len ~ ToothGrowth$dose
anova(lm(anova.model.object))
```