---
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
syl.comp <- read.csv("Syllabus Components.csv")
```

### Workload Expectations

Below you will find:

+ Four assignments that you will complete during the semester,

+ A description of the Project proposal and final project assignment, and

+ General workload expectations.

### Assignment Policy and Procedures
`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Assignment Policy and Procedures")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Assignment Policy and Procedures 2")])`

`r paste0(syl.comp$VAR2[which(syl.comp$VAR1 == "Assignment Policy and Procedures 3")])`

#### Turning in code

You will save your code to your individual-linked dropbox account.

Considerations:

1. You will replace the phrase "LAST.CODE" with your unique last name identifier, the first four characters of your last name. So, if your name is has LAST.CODE 'Smit', then if the assignment asks for the code to be named 'LAST.CODE.initalize.proj.r' you will turn in a file called 'Smit.initalize.proj.r' (Here LAST.CODE has been replaced with 'Smit'). The codes for our class are below.

2. All files submitted must be annotated and have descriptive header information. Annotation and descriptive header information is described in the 'functions' tab on our course website.

3. For assignments that require you to submit multiple files, all of the files will need to run error free and accomplish the tasks for credit to be earned.

4. Every assignment is an Rstudio project. If data are used in the assignments, then the data must be linked through the 'data' directory. If figures are created, then they must be written to the 'figs' directory.

5. You will upload all project files to the proper directory in your linked dropbox.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
require(knitr)
require(kableExtra)
student.list <- read.csv("./Students 2020/student.list.csv")
names(student.list) <- c("First Name",	"Last Name",	"LAST.CODE")
knitr::kable(student.list) %>%
column_spec(3, bold = T, width = "5em") %>%
  kable_styling(bootstrap_options = "striped", full_width = T, position = "center")
```

### Assignment 01 - Working in R

Make a project called 'Assignment 01'.

Create a function named '01.LAST.CODE.initalize.proj.r' that automatically creates a directory structure outlined in (https://nicercode.github.io/blog/2013-04-05-projects/). Save this function into the 'R' directory.

Change your R site profile so that this function loads every time you start R. (https://www.r-bloggers.com/fun-with-rprofile-and-customizing-r-startup/).

Make a new function '01.LAST.CODE.Task.01.r', save this in the R directory. 

This code in '01.LAST.CODE.Task.01.r' will do all of the following: write the entire mtcars data as a csv file to the 'data' directory (mtcars is in the package 'datasets'), import the .csv file, selecting only the first 10 rows only. From this imported and reduced data frame, make a new data frame that includes a column that is the hp per cylinder (divide hp by cyl), name this column hp.per.cyl. Print the data frame to the screen and also save it to the 'data' directory as mtcars.red.RData (it will be saved as file type .RData).

You will graded on whether these files run and return the expected components:

'01.LAST.CODE.initalize.proj.r'

'01.LAST.CODE.Task.01.r'

### Assignment 02 - Figures

Make a project called 'Assignment 02'.

The project will have the sterotyped directory structure described in (https://nicercode.github.io/blog/2013-04-05-projects/) and you will use '01.LAST.CODE.initalize.proj.r' to initialize the project directory (create the direcotry structure).

For each of the excercises in:

'28.2.1 Exercises' (n = 3)

Wrap each of the excercise code in a function: '02.LAST.CODE.fig.01.r', '02.LAST.CODE.fig.02.r', and '02.LAST.CODE.fig.03.r'.

The functions will have at least two function arguments that can be changed by the user to alter the aesthetics or data in the figure.

The function will write the output figure to a .png file in the 'fig' directory in your project. The title of the figures will be:

'02.LAST.CODE.fig.01.png', '02.LAST.CODE.fig.02.png', and '02.LAST.CODE.fig.03.png'

### Assignment 03 - The Tidyverse and Loops

You will make a project called 'Assignment 03'. The project will have the sterotyped directory structure described in (https://nicercode.github.io/blog/2013-04-05-projects/). Use your function '01.LAST.CODE.initalize.proj.r' to accomplish this.

You will practice coding using tidyverse functions found in this link focusing on the Activities in R4DS 05. We will compare the speed of execution of "filter" and "summarize" on a large data frame. We will use with the babynames data from the 'babynames' package. The babynames dataset has historical information about births of babies in the U.S. View a preview of this dataset by typing babynames in to the console. Yes, this dataset contains 1.8 million observations. 

You will create a series of functions that will allow side by side analysis of traditional loops and contemporary tidyverse.

The file '03.LAST.CODE.Task.01.r' will be a script that contain three functions:

The first will be a function that contains the code arranging the babynames tibble in alphabetical order. You will arrange the rows such that the names column is arranged alphabetically. Here you will use tidyverse functionality. The function will be called 'tidy.arrange'.

The second function will arranges the baby names alphabetically but do so without using tidyverse functions. Instead, you will use a loop and base R syntax. This function will be called 'base.arrange'. Recommend identifying unique babynames and looping through these to create a new data frame. There are a lot of ways to accomplish this and I look forward to seeing your creativity.

Both of these functions will return two objects. The first is a data frame of the ordered babynames, the second is the time that it takes the function to exectue. Hint, to get the time use a function that returns the time from computer's internal clock at the start and finish of the function (I use package matlab's functions tic and toc for this, there are other ways). Each function will have one function argument, the input data, babynames.

Now you will write thrird function that calls 'tidy.arrange' and 'base.arrange', this will be called 'arrange.test'. This function will run both tidy.arrange and base.arrange, the intention is to run each function on differnt sized data to understand (using a figure created in this function) how they compare in terms of time to completion. This function will have as an argument the input data used by both functions. The function will return three objects; the times of completion for 'tidy.arrange' and 'base.arrange' and the number of rows in the data frame.

Make a new r script (a .r file) called '03.LAST.CODE.arrange.test.r'. This script will: 'source' (?source) the functions in '03.LAST.CODE.Task.01.r', generate data for plotting (I would recommend to subset the data using the 'seq' function and run arrange.test multiple times), and create a plot that has the number of seconds on the x-axis and the number of rows of the input data on the y-axis. Write this figure to the figs directory as a .png file, '03.LAST.CODE.Task.02.png'.

You will graded on whether these files run. Therefore include the directory structure in your upload to Assignment 03.

### Assignment 04 - Rmarkdown

You will make a project called 'Assignment 04'.

Create a Rmarkdown document (.Rmd) that descibes and documents the motivation and approach, code and output, and figure(s) you created in Assignment 03. 

This Rmarkdown document '04.LAST.CODE.Task.01.Rmd' will pull the functions created in Assignment 03 so note that the figure and function locations (directories in Assignment 03) will need to be consistent with (https://nicercode.github.io/blog/2013-04-05-projects/). This will also be necessary because I will be knitting the submitted .Rmd on my machine. 


### Project

#### Requirement 01: 

You will upload a .Rmd file that I will knit to a .doc and .html file. This document will include your name and the title of your project. You will write approximately 400 words allocated into three sections:

1.) Introduce your project - three sentences.

2.) Describe the computational methods and the data you will use (most of your text here using topic sentences and paragraphs) - data must be larger than 2,000 records and must be publicly available, and 

3.) Describe the deliverables (at least three) that you will accomplish (one sentence per deliverable),

4.) Present the project idea to the class - 10 minute presentation using only material you have provided me in the .Rmd file.

The document will be titled "LAST.CODE.Proj.Prop.Rmd" and be uploaded into the 'doc' directory, located in your personal 'Project' directory. You will meet in person or by phone, with me, to discuss the project scope and objectives prior to submitting the .Rmd file. I will not be available to meet with you after W, April 1, 2020 5 PM. It is your responsiblity to make an appointment with me to discuss your project prior to this date. You will turn this assignment in on W, April 06, 2020 5 PM.  

You will be graded with a 1 or 0 if you have fulfilled all of the requirements: 

1. Meet with me prior to or on W, April 1, 2020 5 PM to discuss the project, 

2. Submit the .Rmd file that can be kitted (no credit if it cannot), and

3. Have fulfilled the formatting and length requirements I document above. 

Requirement 01 is worth 1 point.

#### Requirement 02: 

You will prepare an .Rmd file of your project. I will knit this to an .html file that you will present to the class.

A zero or one will be assigned based on the level of complexity of the analysis and amount of orignial coding and having the .html knit. Only projects that contain every aspect of import, tidy, transform, visualize, model, and communicate can be given credit. The code you provide in your project will be extensive and will document your mastery of the tidyverse, data visualization, and original and interesting analysis. My expectatation is that the coding will be extensive and demonstrate the philosophy of reproducible workflow that we have studied. 

A zero or one will be assigned based on the quality of the talk and the quality of the .html presentation. You will have 10 minutes for your presentation. The visuals and narrative should be interesting, polished, and concise.

The assignment is due: W, May 06, 2020 by 5 PM.

Requirement 02 is worth 2 points.

### General workload expectations

Learning is not a spectator sport. For the engaged learner, mastery of the subject matter covered in this course will require significant coding on your part. To accomplish this you will make directories and projects in your personal directory "RD4DS.01","RD4DS.02", "RD4DS.03", ect. that will include coding from that section. 